References for PloS paper

### 2014-02-14 image_analysis_paper/paper_draft.tex
Upadted draft

@@

@@

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

When a data set is invariant under an underlying symmetry group (such as translations and rotations), we must factor out the underlying symmetries and align the data before doing further analysis.
%
In our example, we will translate and rotate images until they are optimally aligned, and then operate on this aligned set. 
%
There are many ways to align images. 
%
``Template--based'' alignment \cite{...} is one technique, where one selects a specific image or {\em template} as

Angular synchronization calculates the optimal alignments for each point in a data set, given estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.

An important part of angular synchronization is that the symmetry group be {\em compact}. 
%
This is a rather technical point, but all we wish to state is that the group of two--dimensional translations and rotations is {\em not} compact.
%
We can instead map this group to the group of

Let $x_1, \dots, x_n$ denote the data points that we wish to align under some symmetry group.
%
We will consider the case where the symmetry group is the group of $d$-dimensional rotations, $SO(d)$. 
%
Let $R_{ij} \in SO(d)$ denote the rotation that aligns $x_j$ to $x_i$, so that
\begin{equation}
R_{ij} = \argmin_{R \in SO(d)} \|Rx_j - x_i \|^2.
\end{equation}
%
We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
%TODO: Check that this order is correct
%
We would also like to exploit {\em higher-order} consistency relations;
for example, the relationship $R_{ij} R_{jk} \approx R_{ik}$ should hold if the rotation estimates are accurate.
%
Therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.

We would like to solve the optimization problem 
\begin{equation} \label{eqn:angsynch_obj}
\max_{R_1, \dots, R_n \in SO(d)} \left\| \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j \right\|^2.
\end{equation}
%
%TODO: check if a norm is needed
Each term in the objective function will be large large if $R_i$ and $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
%
Therefore, solving \eqref{eqn:angsynch_obj} gives us the set of global alignments which are most consistent with the computed pairwise alignments.

In general, the solution to \eqref{eqn:angsynch_obj} is not easily computed.
%
Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^{d \times d}$.
%
We define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$), with $H_{ij} = R_{ij}$.
%
We can then write the relaxed problem as 
\begin{equation} \label{eqn:angsynch_relax}
\max_{R\in \mathbb{nd \times d}} \| R^T H R \|^2.
\end{equation}
%
The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
%
The $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
%
We would like to note that this formulation also accounts for higher-order relations.
%
For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\| = \| R^T H^2 R \|^2$. 
%
However, the solution to this problem is also given by the top eigenvector of $H$.

Because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{d \times d}$, we must project our approximate solution back to $SO(d)$.
%
The optimal rotation is therefore given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively.

@@

Unlike PCA, diffusion maps (DMAPS) is a nonlinear dimensionality reduction technique. 
%
DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold \cite{coifman2005geometric}.
%
The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
%
This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.

Given data points $x_1, \dots, x_n \in \mathbb{R}^p$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
\begin{equation}
W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon^2} \right)
\end{equation}
and $\epsilon$ is a characteristic distance between data points.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
%
We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{n} W_{ij}$, and the matrix $A = D^{-1} W$. 
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_n$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$. 
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
%
The eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
%
The next few eigenvectors give the ``meaningful'' embedding coordinates for the data.

\subsection{Aligning and Ordering Using Vector diffusion maps}

Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise, and therefore considers all pairwise measurements equally important.
%
However, in system with an underlying symmetry group {\em as well as} a dynamical process, the data points are not identical modulo symmetries and noise, and we would like to place more emphasis on aligning data points which are dynamically close.

Vector diffusion maps (VDM) couples the symmetry-reduction of angular synchronization with the automatic parameterization of diffusion maps \cite{singer2012vector}. 
%
We construct the matrix $A$ as in diffusion maps, and the matrix $H$ as in angular synchronization.
%
We then construct the block matrix $S$, where the blocks of $S$ are defined by
\begin{equation}
S_{ij} = \left\{ \begin{array}{l l} 
A_{ij} H_{ij} & i \ne j \\
0_{d \times d} & i = j
\end{array}
\right.
\end{equation}
%
where $A_{ij}$ is the $(i,j)$ entry of $A$, and $H_{ij}$ is the $(i,j)$ block of $H$.

We compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
%
As in angular synchronization, the first ``block'' eigenvector, i.e. the first $d$ eigenvectors concatenated into a $nd \times d$ matrix, gives the (relaxed) optimal rotations. 
%
However, the eigenvectors now also give us embedding coordinates.
%
Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$ (so that $\phi_j(i) \in \mathbb{R}^d$). 
%
The embedding coordinates of $x_i$ are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.
